---
title: "East Maui Irrigation Model Report"
author: "Trutta Environmental Solutions, LLC"
output: word_document
---

```{r setup, include=FALSE, echo=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(eastMaui)
library(tidyverse)
library(gridExtra)
library(knitr)

percent <- function(x, digits = 2, format = "f", ...) {
  paste0(formatC(100 * x, format = format, digits = digits, ...), "%")
}

```

## Methods

**Add text here to explain methods**


```{r input generation}
# 
watersheds = inputs
watersheds = watersheds[ ,-16]

watershedsWaterInput = watersheds[ ,c(1,2,7)]
waterInput.mat= as.matrix(watershedsWaterInput)

nodes = eastMaui::nodes
nodescurr = nodes[ ,c(1, 9:18)]
nodes00 = nodescurr[1, ]
nodes2 = nodescurr[-1, ]
nodescurr = rbind(nodes2, nodes00)
nodes.mat = as.matrix(nodescurr[ ,-2])

# ----------------------------------------------------------------------------------
# Create nodes.mat for all values = 1

nodes1 = eastMaui::nodes
nodes1 = nodes1[ ,c(1, 9:18)]
nodes1[ , c(3:12)] <- 1
# View(nodes1)

nodes1.mat = as.matrix(nodes1[ ,-2])
row.names(nodes1.mat)<- unlist(nodes1[ ,1])
nodes1.mat[ , c(3:11)] = as.numeric(nodes1.mat[ , c(3:11)] )
# View(nodes1.mat)
# ----------------------------------------------------------------------------------
# Create nodes.mat for all values = 0

nodes0 = eastMaui::nodes
nodes0 = nodes0[ ,c(1, 8:18)]
nodes0[which(nodes0$NodeType == "MinorDiversion"), c(3:12)] <- 0
nodes0[which(nodes0$NodeType == "MajorDiversion"), c(3:12)] <- 0
nodes0.mat = as.matrix(nodes0[ ,-2])
row.names(nodes0.mat)<-unlist(nodes0[ ,1])
nodes0.mat[ , c(3:11)] = as.numeric(nodes0.mat[ , c(3:11)] )
# View(nodes0.mat)


```


### Habitat calculations
**Explanatory text about hab calcs**

```{r Scenarios - habitat}
# Run convertMatrix.R
matrices = convertMatrix(nodes, watersheds)

# Run scenarios

scenario.current = basin.fun(matrices$nodes, matrices$inputs)

scenario.1 = basin.fun(nodes1.mat, matrices$inputs)

scenario.0 = basin.fun(nodes0.mat, matrices$inputs)

# --------------------------------------------------------------------------------------------------
# append outputs

scenario.c = as.data.frame(t(scenario.current))
scenario.1t = as.data.frame(t(scenario.1))
scenario.0t = as.data.frame(t(scenario.0))

scenario.c = tibble::rownames_to_column(scenario.c, var = "rowname")

colnames(scenario.1t) = paste("1", colnames(scenario.1t), sep = "_")
colnames(scenario.0t) = paste("0", colnames(scenario.0t), sep = "_")

scenario.test = bind_cols(scenario.c, scenario.0t) %>%
  bind_cols(scenario.1t) %>%
  separate(rowname, into = c("E", "Wshed"), sep = "^([E])")

# match basinID to watershed

colnames(scenario.test)[1] = 'Wshed'
colnames(scenario.test)[2] = "BASINID"

scenario.test$BASINID = as.numeric(scenario.test$BASINID)
scenario.test$Wshed = as.numeric(scenario.test$Wshed)
wsheds = watersheds[ ,c(1,2)]

scenario.test = left_join(scenario.test, wsheds, by = "BASINID")
wsheds = scenario.test[ , 27]
scenario.test = scenario.test[ , -27]
scenario.test$Wshed = wsheds
# write_xlsx(scenario.test, "scenario_test.xlsx")

# Watershed totals for each scenario
for(i in 1:nrow(scenario.test)){
  scenario.test$Wshed.total[i] = sum(scenario.test[i, c(3:10)])
}
for(i in 1:nrow(scenario.test)){
  scenario.test$Wshed.total1[i] = sum(scenario.test[i ,c(19:26)])
}
for(i in 1:nrow(scenario.test)){
  scenario.test$Wshed.total0[i] = sum(scenario.test[i ,c(11:18)])
}

# generate new ID variables
scenario.test$IDx1 = sample(1:3, nrow(scenario.test), replace = TRUE)
scenario.test$IDx2 = sample(1:5, nrow(scenario.test), replace = TRUE)

scenario.test = scenario.test[ , c(1,2,30,31, 3:29)]
```

### Water diversion calculations
**Explanatory text about water calcs**


```{r Scenarios - diversions}

# waterInput = waterInput.mat

scenario.currentw = diversion.fun(nodes.mat, waterInput.mat)

scenario.1w = diversion.fun(nodes1.mat, waterInput.mat)

scenario.0w = water.fun(nodes0.mat, waterInput.mat)

# --------------------------------------------------------------------------------------------------
# append outputs
scenario.currentw = scenario.currentw[1, ]
scenario.1w = scenario.1w[1,]
scenario.0w = scenario.0w[1,]

scenario.cw = as.data.frame(t(scenario.currentw))
scenario.1tw = as.data.frame(t(scenario.1w))
scenario.0tw = as.data.frame(t(scenario.0w))

scenario.cw = tibble::rownames_to_column(scenario.cw, var = "rowname")
colnames(scenario.cw)[2] = "MixedScen"
scenario.0tw = rownames_to_column(scenario.0tw)
scenario.1tw = rownames_to_column(scenario.1tw, var = "rowname")
colnames(scenario.1tw) = c("rowname", "All_1")
colnames(scenario.0tw) = c("rowname","All_0")

scenario.testw = merge(scenario.cw, scenario.0tw, by = "rowname" ) %>%
  left_join(scenario.1tw) %>%
  separate(rowname, into = c("E", "Wshed"), sep = "^([E])")

# match basinID to watershed

colnames(scenario.testw)[1] = 'Wshed'
colnames(scenario.testw)[2] = "BASINID"
colnames(scenario.testw)[3] = "mixed"
scenario.testw$BASINID = as.numeric(scenario.testw$BASINID)
scenario.testw$Wshed = as.numeric(scenario.testw$Wshed)
wsheds = watersheds[ ,c(1,2)]

scenario.testw = left_join(scenario.testw, wsheds, by = "BASINID")
scenario.testw = scenario.testw[ ,-1]
scenario.testw = scenario.testw[ ,c(5,1:4)]

# generate new ID variables
scenario.testw$IDx1 = sample(1:3, nrow(scenario.testw), replace = TRUE)
scenario.testw$IDx2 = sample(1:5, nrow(scenario.testw), replace = TRUE)
scenario.testw = scenario.testw[ ,c(1:2,6:7, 3:5)]



```


## Output

### Grouping Variable = Watersheds
#### Habitat results
```{r Output habitat by watershed}
# output = summary of habitat by watershed and by species
WshedID = unique(scenario.test$Wshed)
WshedCount = c(1:n_distinct(WshedID))
summaries = tibble(WshedID, WshedCount)
species.abbvs = c("AB", "AS", "ES", "LC", "MG", "NG", "SH", "SS")

output = scenario.test %>%
  group_by(Wshed) %>%      #LINE to change GROUP ID var
  summarise_at(vars(c(4:27)), sum)

label = output[, 1]
All0 = round(output[ ,c(10:17)], 2)
# original = round(original, 2)
percentChangeMixed = ((output[ ,c(2:9)]- output[ ,c(10:17)])/output[ ,c(10:17)])*100
percentChangeAll1 = ((output[ ,c(18:25)]- output[ ,c(10:17)])/output[ ,c(10:17)])*100

scenario.changes = bind_cols(label,All0,percentChangeMixed, percentChangeAll1)
scenario.changes[is.na(scenario.changes)]<- 0

# *******************************************
# summaries by Wshed
Wshed = c(1:n_distinct(scenario.changes))
totalHab.All0 = rowSums(All0)
totalHab.All0[is.na(totalHab.All0)]<- 0
totalHab.mixed = rowSums(output[ ,c(2:9)])
totalHab.mixed[is.na(totalHab.mixed)]<- 0
totalHab.All1 = rowSums(output[ ,c(18:25)])
totalHab.All1[is.na(totalHab.All1)]<- 0
Wshed.out = data.frame(Wshed, totalHab.All0, totalHab.mixed, totalHab.All1)

Wshed.out$percentDelta.mixed = ((Wshed.out[, 3]-Wshed.out[ ,2])/Wshed.out[,2])*100
Wshed.out$percentDelta.mixed[is.na(Wshed.out$percentDelta.mixed)]<- 0
Wshed.out$percentDelta1 = ((Wshed.out[, 4]-Wshed.out[ ,2])/Wshed.out[,2])*100
Wshed.out$percentDelta1[is.na(Wshed.out$percentDelta1)]<- 0
Wshed.out$percentDelta.mixed[is.infinite(Wshed.out$percentDelta.mixed)]<- 0
Wshed.out$percentDelta1[is.infinite(Wshed.out$percentDelta1)]<- 0

Wshed.long = gather(Wshed.out, `percentDelta.mixed`, `percentDelta1`,  key = "NodeValue", value = "PercentChange")
Wshed.long$PercentChange[is.infinite(Wshed.long$PercentChange)]<- 0
Wshed.long$lPercentChange = log(Wshed.long$PercentChange)
Wshed.long$lPercentChange[is.infinite(Wshed.long$lPercentChange)]<- 0
Wshed.long$lPercentChange[is.infinite(Wshed.long$lPercentChange)]<- 0

ggplot(Wshed.out, aes(x = Wshed, y = log(`totalHab.All0`))) +
  geom_bar(stat = 'identity') +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(y = "log(All 0)") +
  theme_classic()+
  ggtitle("Watershed habitat summary - Baseline")

ggplot(Wshed.long, aes(x = Wshed, y = lPercentChange, fill = NodeValue)) +
  geom_bar( stat = 'identity', position = 'dodge') +
  ylim(0, 1.2*max(Wshed.long$lPercentChange)) +
  labs(y = "log(PercentChange)") +
  theme_classic() +
  ggtitle("Watershed Habitat Changes under New Scenarios")

# generate summary table
colnames(Wshed.out)[2:6] = c("Total Habitat 0", "Total Habitat Mixed", "Total Habitat 1", "Percent Change Mixed", "Percent Change 1")

Wshed.out$`Percent Change Mixed`=percent(Wshed.out$`Percent Change Mixed`)

Wshed.out$`Percent Change 1`=percent(Wshed.out$`Percent Change 1`)

Wshed.tab = kable(Wshed.out, digits = 1, align = 'c',
                  caption = "Summary of Habitat Changes by Watershed")
Wshed.tab
```

**Space for interpretation of preceding plots and table**.

#### Water Diversion results - Water shunted to diversions grouped by watershed
```{r Output water by watershed}

# output = summary of water by watershed

outputw = scenario.testw %>%
  group_by(WshedID) %>%      #LINE to change GROUP ID var
  summarise_at(vars(mixed:All_1), sum)

# *******************************************
# calculate percent change from output
label = outputw[ ,1]
All0 = outputw[ ,3]
colnames(All0)[1] = "All0"
percentChangeMixed = ((outputw[ ,c(2)]- outputw[ ,c(3)])/outputw[ ,c(3)])*100
percentChangeAll1 = ((outputw[ ,c(4)]- outputw[ ,c(3)])/outputw[ ,c(3)])*100

scenario.changesw = bind_cols(label,All0,percentChangeMixed, percentChangeAll1)
scenario.changesw[is.na(scenario.changesw)]<- 0
colnames(scenario.changesw)[3]<- c("pChangeMixed")
colnames(scenario.changesw)[4]<- c("pChangeAll1")

# *******************************************
# summaries by watersheds
Wshed = c(1:n_distinct(scenario.changesw))
totalDiv.All0 = rowSums(All0)
totalDiv.All0[is.na(totalDiv.All0)]<- 0
totalDiv.mixed = rowSums(outputw[ ,c(2)])
totalDiv.mixed[is.na(totalDiv.mixed)]<- 0
totalDiv.All1 = rowSums(outputw[ ,c(4)])
totalDiv.All1[is.na(totalDiv.All1)]<- 0
Wshed.outw = data.frame(Wshed, totalDiv.All0, totalDiv.mixed, totalDiv.All1)

Wshed.outw$percentDelta.Mixed = ((Wshed.outw[, 3]-Wshed.outw[ ,2])/Wshed.outw[,2])*100
Wshed.outw$percentDelta.Mixed[is.na(Wshed.outw$percentDelta.Mixed)]<- 0
Wshed.outw$percentDelta1 = ((Wshed.outw[, 4]-Wshed.outw[ ,2])/Wshed.outw[,2])*100
Wshed.outw$percentDelta1[is.na(Wshed.outw$percentDelta1)]<- 0
Wshed.outw$percentDelta.Mixed[is.infinite(Wshed.outw$percentDelta.Mixed)]<- 0
Wshed.outw$percentDelta1[is.infinite(Wshed.outw$percentDelta1)]<- 0
# reshape data to plot percent delta

Wshed.outw.long = gather(Wshed.outw, `percentDelta.Mixed`, `percentDelta1`,  key = "NodeValue", value = "PercentChange")
Wshed.outw.long$PercentChange[is.infinite(Wshed.outw.long$PercentChange)]<- 0
Wshed.outw.long$lPercentChange = log(Wshed.outw.long$PercentChange)
Wshed.outw.long$lPercentChange[is.infinite(Wshed.outw.long$lPercentChange)]<- 0
Wshed.outw.long$lPercentChange[is.infinite(Wshed.outw.long$lPercentChange)]<- 0


# barplots
ggplot(Wshed.outw, aes(x = Wshed, y = (totalDiv.All0))) +
  geom_bar(stat = 'identity') +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(y = "All 0") +
  theme_classic()+
  ggtitle("Watershed Diversions - Baseline")

ggplot(Wshed.outw.long, aes(x = Wshed, y = PercentChange, fill = NodeValue)) +
  geom_bar( stat = 'identity', position = 'dodge') +
  ylim(0, 1.2*max(Wshed.outw.long$lPercentChange)) +
  labs(y = "PercentChange") +
  theme_classic() +
  ggtitle("Watershed Diversions - New Scenarios")

# summary table
colnames(Wshed.outw)[2:6] = c("Total Diverted - All 0", "Total Diverted - Mixed", "Total Diverted - All 1", "Percent Change - Mixed", "Percent Change - All 1")

Wshed.outw$`Percent Change - Mixed`= percent(Wshed.outw$`Percent Change - Mixed`)
Wshed.outw$`Percent Change - All 1`= percent(Wshed.outw$`Percent Change - All 1`)

Wshed.outw.tab = kable(Wshed.outw, digits = 1, align = 'c',
                       caption = "Summary of Water Diversions by Watershed")
Wshed.outw.tab

```


**Space for interpretation of preceding plots and table**.

#### Habitat results by species, grouped by Watershed
```{r Habitat output by species}

species.abbvs = c("AB", "AS", "ES", "LC", "MG", "NG", "SH", "SS")

AB = scenario.changes[ ,c(1,2,10,18)]
AS = scenario.changes[ ,c(1,3,11,19)]
ES = scenario.changes[ ,c(1,4,12,20)]
LC = scenario.changes[ ,c(1,5,13,21)]
MG = scenario.changes[ ,c(1,6,14,22)]
NG = scenario.changes[ ,c(1,7,15,23)]
SH = scenario.changes[ ,c(1,8,16,24)]
SS = scenario.changes[ ,c(1,9,17,25)]
species.list = list(AB,AS,ES,LC,MG,NG,SH,SS)
species.plots = list()

# list of barplots
for(i in 1:length(species.list)){
  colnames(species.list[[i]])<- c("Wshed", "All 0", "Mixed", "All 1")
  
  species.list[[i]][,5] = ((species.list[[i]][ ,2]-species.list[[i]][,3])/species.list[[i]][,3])*100
  species.list[[i]][,6] = ((species.list[[i]][ ,4]-species.list[[i]][,3])/species.list[[i]][,3])*100
  colnames(species.list[[i]])[c(5,6)]<-c("PercentChangeMixed", "PercentChangeAll1")
  species.list[[i]][is.na(species.list[[i]])]<- 0
  
  dat.long = gather(species.list[[i]], "PercentChangeMixed", "PercentChangeAll1", key = "NodeValue", value = "PercentChange")
  dat.long$lPercentChange = log(dat.long$PercentChange)
  dat.long$lPercentChange[is.na(dat.long$lPercentChange)]<- 0
  dat.long$lPercentChange[is.infinite(dat.long$lPercentChange)]<- 0
  
  
  g.sp = ggplot(dat.long, aes(x = Wshed, y = lPercentChange, fill = NodeValue )) +
    geom_bar( stat = 'identity', position = 'dodge') +
    theme_classic() +
    ggtitle(paste0("Species: ", species.abbvs[i], "; Grouped by: ", colnames(scenario.changes)[1], sep = " "))
  species.plots[[i]] <- g.sp

}

# list of tables
species.tabs = list()
for(i in 1:length(species.list)){
  species.list[[i]]$PercentChangeMixed = percent(species.list[[i]]$PercentChangeMixed)
  species.list[[i]]$PercentChangeAll1 = percent(species.list[[i]]$PercentChangeAll1)
  df = as.data.frame(species.list[[i]])
  df$Mixed[is.infinite(df$Mixed)]<- 0
  df$`All 1`[is.infinite(df$`All 1`)]<- 0
  # summary tables for each species
  IDvar.tab = kable(df, digits = 1, align = 'c')
  
  species.tabs[[i]]<- IDvar.tab
  
}

species.plots[[1]]
species.tabs[[1]]

species.plots[[2]]
species.tabs[[2]]

species.plots[[3]]
species.tabs[[3]]

species.plots[[4]]
species.tabs[[4]]

species.plots[[5]]
species.tabs[[5]]

species.plots[[6]]
species.tabs[[6]]

species.plots[[7]]
species.tabs[[7]]

species.plots[[8]]
species.tabs[[8]]

```

**Space for interpretation of preceding plots and tables**.
