---
title: "East Maui Irrigation Model Report"
author: "Trutta Environmental Solutions, LLC"
output: word_document
---

```{r setup, include=FALSE, echo=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(eastMaui)
library(tidyverse)
library(gridExtra)
library(knitr)
library(readxl)
library(openxlsx)


inpath = "~/Programming/Trutta/HSHEP/EMaui/EMpackage/eastMaui/rawData"
# required functions
percent <- function(x, digits = 2, format = "f", ...) {
  paste0(formatC(100 * x, format = format, digits = digits, ...), "%")
}
allsp.hab = function(x){
  y = (1-(0.0000681*((100-x)^2) - 0.00032*(100-x)))
  return(y)  
}
ab.hab = function(x){
  y = (1-(0.00004678*((100-x)^2) + 0.000313*(100-x)))
  return(y)
}

# r input generation
# 
watersheds = read_excel(paste(inpath, "/watersheds.xlsx", sep = ""))

# Water calculations input
watershedsWaterInput = watersheds[ ,c(1,2,8)]
waterInput.mat= as.matrix(watershedsWaterInput)

# Habitat calculations input matrix
habitatInput.mat = data.matrix(watersheds)
habitatInput.mat = habitatInput.mat[, -6]


# nodes.compare = eastMaui::nodes1

nodes = read_excel(paste(inpath, "/DiversionInputMix5.xlsx", sep = ""))
nodes = nodes[ ,-c(2:7)]
nodes0 = read_excel(paste(inpath, "/nodes_fulldiv.xlsx", sep = ""))
nodes0 = nodes0[ ,-c(2:7)]
nodes1 = read_excel(paste(inpath, "/nodes_nodiv.xlsx", sep = ""))
nodes1 = nodes1[ ,-c(2:7)]
nodes.mat = data.matrix(nodes)
nodes0.mat = data.matrix(nodes0)
nodes1.mat = data.matrix(nodes1)

# calc upstream/downstream effect for each node
# for habitat calculations
nodesh = nodes
nodes0h = nodes0
nodes1h = nodes1

nodesh$dsImpact = nodesh$DnEntrain*nodesh$DnBarrier*nodesh$DnHab*nodesh$DnOther
nodes0h$dsImpact = nodes0h$DnEntrain*nodes0h$DnBarrier*nodes0h$DnHab*nodes0h$DnOther
nodes1h$dsImpact = nodes1h$DnEntrain*nodes1h$DnBarrier*nodes1h$DnHab*nodes1h$DnOther

nodesh$usImpact = nodesh$UpEntrain*nodesh$UpBarrier*nodesh$UpOther*nodesh$UpHab
nodes0h$usImpact = nodes0h$UpEntrain*nodes0h$UpBarrier*nodes0h$UpOther*nodes0h$UpHab
nodes1h$usImpact = nodes1h$UpEntrain*nodes1h$UpBarrier*nodes1h$UpOther*nodes1h$UpHab

nodesh$passage = nodesh$dsImpact*nodesh$usImpact
nodes0h$passage = nodes0h$dsImpact*nodes0h$usImpact
nodes1h$passage = nodes1h$dsImpact*nodes1h$usImpact

nodesh.mat = data.frame(nodesh)
nodes0h.mat = data.frame(nodes0h)
nodes1h.mat = data.frame(nodes1h)

# --------------------------------------------------
# Water diversion calculations
waterInput = waterInput.mat

sets = c("inStream", "inWshed", "inDitch", "springs", "sinks", "WshedCheck")

outpath = "~/Programming/Trutta/HSHEP/EMaui/EMpackage/eastMaui/outputs/"
# ---------------------------------------------------
# Run function w/ different scenarios

test1 = basinWater.fun(nodes.mat, waterInput.mat)
test2 = basinWater.fun(nodes0.mat, waterInput.mat)
test3 = basinWater.fun(nodes1.mat, waterInput.mat)

# ---------------------------------------------------
instream1 = as.data.frame(test1$inStream)
instream2 = as.data.frame(test2$inStream)
instream3 = as.data.frame(test3$inStream)

instream = bind_rows(instream1, instream2, instream3)
instream = t(instream)
colnames(instream)<-c("mixed", "all0", "all1")
instream = as.data.frame(instream) %>%
  rownames_to_column(var = "basinID") %>%
  separate(basinID, into = c("E", "Basin"), sep = "^([E])")
instream$Basin = as.numeric(instream$Basin)
instream = instream[-53,]
instream[75, 2] = 77

add206 = c(0,206,0,0,0)
names(add206) = c("E", "Basin", "mixed", "all0", "all1")
instream = rbind(instream, add206)

add271 = c(0,271,0,0,0)
names(add271) = c("E", "Basin", "mixed", "all0", "all1")
instream = rbind(instream, add271)
add281 = c(0,281,0,0,0)
names(add281) = c("E", "Basin", "mixed", "all0", "all1")
instream = rbind(instream, add281)

# ---- habitat calculation inputs----------------------
instreamh = instream[order(instream$Basin), -1]
instreamh = remove_rownames(instreamh) 


instreamh.mat = data.matrix(instreamh)

instreamh.nat = instreamh.mat[ ,4]

instreamh.0 = instreamh.mat[ ,3]
instreamh.mix = instreamh.mat[ ,2]

waterImpactedHab.nat = (instreamh.nat/instreamh.nat)*100
waterImpactedHab.0 = (instreamh.0/instreamh.nat)*100
waterImpactedHab.mix = (instreamh.mix/instreamh.nat)*100

waterImpactedHab.nat[is.na(waterImpactedHab.nat)]<- 0
waterImpactedHab.0[is.na(waterImpactedHab.0)]<- 0
waterImpactedHab.mix[is.na(waterImpactedHab.mix)]<- 0

# waterImpactedHab = waterImpactedHab.mix


# -----------------------------------------------------
inwshed1 = as.data.frame(test1$inWshed)
inwshed2 = as.data.frame(test2$inWshed)
inwshed3 = as.data.frame(test3$inWshed)

inwshed = bind_rows(inwshed1, inwshed2, inwshed3)
inwshed = t(inwshed)
colnames(inwshed)<-c("mixed", "all0", "all1")
inwshed = as.data.frame(inwshed) %>%
  rownames_to_column(var = "wshedID") %>%
  separate(wshedID, into= c("w", "Wshed"), sep = "^([w])")
inwshed$Wshed = as.numeric(inwshed$Wshed)
inwshed = inwshed[, -1]
# -----------------------------------------------------
inditch1 = as.data.frame(test1$inDitch)
inditch2 = as.data.frame(test2$inDitch)
inditch3 = as.data.frame(test3$inDitch)

inditch1[which(inditch1<0)]<- 0
inditch2[which(inditch2<0)]<- 0
inditch3[which(inditch3<0)]<- 0

inditch = bind_rows(inditch1, inditch2, inditch3)
inditch = t(inditch)
colnames(inditch)<-c("mixed", "all0", "all1")

inditch = as.data.frame(inditch) %>%
  rownames_to_column(var = "ditchID") %>%
  separate(ditchID, into = c("n", "NodeID"), sep = "^([n])")
inditch$NodeID = as.numeric(inditch$NodeID)
inditch = inditch[ , -1]

# convert to mgd - millions of gallons per day
inditch$mixed.mgd = inditch$mixed*0.646317
inditch$all0.mgd = inditch$all0*0.646317
inditch$all1.mgd = inditch$all1*0.646317
# ---------------------------------------------------
wshedcheck1 = as.data.frame(test1$WshedCheck)
wshedcheck2 = as.data.frame(test2$WshedCheck)
wshedcheck3 = as.data.frame(test3$WshedCheck)

check = bind_rows(wshedcheck1, wshedcheck2, wshedcheck3)
check = t(check)
colnames(check)<- c("mixed", "all0", "all1")

check = as.data.frame(check) %>%
  rownames_to_column(var = "wshed") %>%
  separate(wshed, into = c("w", "Wshed"), sep = "^([w])")
check$Wshed = as.numeric(check$Wshed)
check = check[, -1]

# ----------------------------------------------

springs1 = as.data.frame(test1$springs)
springs2 = as.data.frame(test2$springs)
springs3 = as.data.frame(test3$springs)

springs = bind_rows(springs1, springs2, springs3)
springs = t(springs)
colnames(springs)<- c("mixed", "all0", "all1")

springs = as.data.frame(springs) %>%
  rownames_to_column(var = "springID") %>%
  separate(springID, into = c("E", "spring_ID"), sep = "^([E])")
springs = springs[ , -1]

# ----------------------------------------------

sinks1 = as.data.frame(test1$sinks)
sinks2 = as.data.frame(test2$sinks)
sinks3 = as.data.frame(test3$sinks)

sinks = bind_rows(sinks1, sinks2, sinks3)
sinks = t(sinks)

colnames(sinks)<- c("mixed", "all0", "all1")

sinks = as.data.frame(sinks) %>%
  rownames_to_column(var = "sinkID") %>%
  separate(sinkID, into = c("E", "sink_ID"), sep = "^([E])")
sinks = sinks[ , -1]
# ---------------------------------------------------
# prep inputs for export to final workbook
waterOutput = watersheds[ ,-c(17:24)]
waterOutput = waterOutput[ ,c(1:2, 17:21, 3:16)]

nodespath = "~/Programming/Trutta/HSHEP/EMaui/EMpackage/eastMaui/rawData"
nodesOutput = read_excel(paste(nodespath, "/nodes_for_output.xlsx", sep = ""))

dsNodes = read_excel(paste(nodespath, "/BasinsAllGroups.xlsx", sep = ""))
dsNodes = dsNodes[ ,c(1,2,25)]

# ---------------------------------------------
# results
output_list = list("watershedsInput" = waterOutput,
                   "diversionsInput" = nodesOutput,
                   "WshedCheck" = check,
                   "inStream" = instream,
                   "inWshed"= inwshed,
                   "inDitch" = inditch,
                   "springs" = springs,
                   "sinks" = sinks)
file = paste(outpath, "water_outputs.0.5.xlsx", sep = "")
Sys.setenv("R_ZIPCMD" = "C:/RBuildTools/3.4/bin/zip.exe")
write.xlsx(output_list, file = file)
# ---------------------------------------------
# water modeling key
waterOutput %>% distinct( WshedID, Wshed_Name ) -> wshedNames

colnames(nodesOutput)[1]<- "NodeID"
nodesInfo = nodesOutput[, c(1,4,6,7)]
inditch %>% left_join(nodesInfo, by = "NodeID") -> nodesOutput0
nodesOutput0 = nodesOutput0[which(nodesOutput0$DiversionID %in% c(1:10)), ]
# writexl::write_xlsx(nodesOutput0, path = paste(path, "/nodes_output_summ.1.0.xlsx", sep = ""))

nodesOutput0 %>%
  group_by(DiversionID) %>%
  summarise_at(vars(c(2:4)), sum) -> summByDiversion

leasesByID = waterOutput[ ,c(2,4,7)]
colnames(dsNodes)[3]<- "NodeID"
leases = left_join(dsNodes, leasesByID, by = "BASINID") %>%
  left_join(inditch, by = "NodeID") %>%
  filter(NodeID<45 | NodeID>54) %>%
  group_by(Lease_ID)
leases = leases[complete.cases(leases), ]
leases = leases %>% summarise_at(vars(c(5:7)), sum)

taro = left_join(dsNodes, leasesByID, by = "BASINID") %>%
  left_join(inditch, by = "NodeID") %>%
  filter(NodeID<45 | NodeID>54) %>%
  group_by(Taro_ID)
taro = taro[complete.cases(taro),]
taro = taro %>% summarise_at(vars(c(5:7)), sum)

key_list = list("Watersheds" = wshedNames,
                "Diversion" = summByDiversion,
                "leases" = leases,
                "taro" = taro)
file = paste(outpath, "water_outputs_key.1.1.xlsx", sep = "")
Sys.setenv("R_ZIPCMD" = "C:/RBuildTools/3.4/bin/zip.exe")
openxlsx::write.xlsx(key_list, file = file)
# --------------------------------------------
# --------------------------------------------
# Habitat Modeling
htest = basin.fun(nodesh.mat, habitatInput.mat, waterImpactedHab.mix)
htest1 = basin.fun(nodes1h.mat, habitatInput.mat, waterImpactedHab.nat)
htest0 = basin.fun(nodes0h.mat, habitatInput.mat, waterImpactedHab.0)


AllSpHab = as.data.frame(htest$WsedHabAllSp)
AllSpHab1 = as.data.frame(htest1$WsedHabAllSp)
AllSpHab0 = as.data.frame(htest0$WsedHabAllSp)
AllSpHabm = as.tibble(t(AllSpHab))
AllSpHab1 = as.tibble(t(AllSpHab1))
AllSpHab0 = as.tibble(t(AllSpHab0))

ABspHab = as.tibble(htest$WshedHabAB)
ABspHab1 = as.tibble(htest1$WshedHabAB)
ABspHab0 = as.tibble(htest0$WshedHabAB)
ABspHabm = as.tibble(t(ABspHab))
ABspHab1 = as.tibble(t(ABspHab1))
ABspHab0 = as.tibble(t(ABspHab0))
colnames(ABspHabm) = "AB_sum"
colnames(ABspHab1) = "AB_sum"
colnames(ABspHab0) = "AB_sum"

AllSpHab.mix = bind_cols(AllSpHabm, ABspHabm)
AllSpHab.mix$total = rowSums(AllSpHab.mix)
AllSpHab.nat = bind_cols(AllSpHab1, ABspHab1) 
AllSpHab.nat$total = rowSums(AllSpHab.nat)
AllSpHab.0 = bind_cols(AllSpHab0, ABspHab0)
AllSpHab.0$total = rowSums(AllSpHab.0)


summWshed = c(1:45)
natural.total = tibble(summWshed,AllSpHab.nat$total)
colnames(natural.total)<-c("WshedID", "total.habitat")
mix.total = tibble(summWshed,AllSpHab.mix$total)
colnames(mix.total)<-c("WshedID", "total.habitat")
fullDiv.total = tibble(summWshed, AllSpHab.0$total)
colnames(fullDiv.total)<-c("WshedID", "total.habitat")

addDitches = read_excel('~/Programming/Trutta/HSHEP/Emaui/fixNodesDitches.xlsx', sheet = 1)
colnames(addDitches)<- c("WshedID", "BASINID", "NodeID")
basinDivs = left_join(addDitches, inditch, by = "NodeID" )
basinDivs[is.na(basinDivs)]<- 0
basinDivs[6, c(4:6)]= 0
natural.diversion = basinDivs %>%
  group_by(WshedID) %>%
  summarise_at(vars("all1.mgd"), sum)
mixed.diversion = basinDivs %>%
  group_by(WshedID) %>%
  summarise_at(vars("mixed.mgd"), sum)
full.diversion = basinDivs %>%
  group_by(WshedID) %>%
  summarise_at(vars("all0.mgd"), sum)

natural = left_join(natural.total,natural.diversion, by = "WshedID")
colnames(natural)[c(2,3)]<- c("Natural.Habitat.Units", "no.div.mgd")
mixed = left_join(mix.total, mixed.diversion, by = "WshedID")
colnames(mixed)[c(2,3)]<- c("Mixed.Habitat.Units", "mixed.div.mgd")
full = left_join(fullDiv.total, full.diversion, by = "WshedID")
colnames(full)[c(2,3)]<- c("Full.Diversion.Habitat.Units", "full.div.mgd")

summ = left_join(natural, mixed, by = "WshedID") %>% left_join(full, by = "WshedID")

hab_list = list("Summary" = summ,
                "AllSpeciesHabitat.nat" = AllSpHab.nat,
                "AllSpeciesHabitat.mix" = AllSpHab.mix,
                "AllSpeciesHabitat.0" = AllSpHab.0)
file = paste(outpath, "habitat_outputs.1.1.xlsx", sep = "")
Sys.setenv("R_ZIPCMD" = "C:/RBuildTools/3.4/bin/zip.exe")
openxlsx::write.xlsx(hab_list, file = file)
```






### Amount of water present in each basin
```{r Water Output by Diversion ID}
# Water in Basins
groups = watersheds
match1 = groups[ ,c(1:9)]
scenario.currentw = as.data.frame(basinWater.fun(nodes.mat, waterInput.mat))

scenario.1w = as.data.frame(basinWater.fun(nodes1.mat, waterInput.mat))

scenario.0w = as.data.frame(basinWater.fun(nodes0.mat, waterInput.mat))

# --------------------------------------------------------------------------------------------------
# append outputs
scenario.currentw = scenario.currentw[1, ]
scenario.1w = scenario.1w[1,]
scenario.0w = scenario.0w[1,]

scenario.cw = as.data.frame(t(scenario.currentw))
scenario.1tw = as.data.frame(t(scenario.1w))
scenario.0tw = as.data.frame(t(scenario.0w))

scenario.cw = tibble::rownames_to_column(scenario.cw, var = "rowname")
colnames(scenario.cw)[2] = "MixedScen"
scenario.0tw = rownames_to_column(scenario.0tw)
scenario.1tw = rownames_to_column(scenario.1tw, var = "rowname")
colnames(scenario.1tw) = c("rowname", "All_1")
colnames(scenario.0tw) = c("rowname","All_0")

scenario.testw = merge(scenario.cw, scenario.0tw, by = "rowname" ) %>%
  left_join(scenario.1tw) %>%
  separate(rowname, into = c("E", "Wshed"), sep = "^([E])")

# match basinID to watershed

colnames(scenario.testw)[1] = 'Wshed'
colnames(scenario.testw)[2] = "BASINID"
colnames(scenario.testw)[3] = "mixed"
scenario.testw$BASINID = as.numeric(scenario.testw$BASINID)
scenario.testw$Wshed = as.numeric(scenario.testw$Wshed)
wsheds = watersheds[ ,c(1,2)]

scenario.testw = left_join(scenario.testw, wsheds, by = "BASINID")
scenario.testw = scenario.testw[ ,-1]
scenario.testw = scenario.testw[ ,c(5,1:4)]

basins = scenario.testw[ ,c(2:5)]
datw = left_join(match1, basins, by = "BASINID")
datw$BFQ = watersheds$BFQ95

# bf% - all water in basins/ all_1
writexl::write_xlsx(datw, paste(path, "/WaterInBasinsv0.6.2.xlsx", sep = ""))

datj = watersheds[ ,c(2,9:16)]
# 
# # calculate X for input to functions below
# dat.sp = left_join(datw, datj, by = "BASINID") 
# dat.sp$divert.mixed = ((dat.sp$mixed-dat.sp$BFQ)/dat.sp$BFQ)*100
# dat.sp$divert.0 = ((dat.sp$All_0-dat.sp$BFQ)/dat.sp$BFQ)*100
# dat.sp$divert.1 = ((dat.sp$All_1-dat.sp$BFQ)/dat.sp$BFQ)*100
# dat.sp$divert.mixed1 = dat.sp$mixed/dat.sp$BFQ
# dat.sp$divert.01 = dat.sp$All_0/dat.sp$BFQ
# dat.sp$divert.11 = dat.sp$All_1/dat.sp$BFQ

# calculate water mediated effect on habitat for each basin
allsp.hab = function(x){
  y = (1-(0.0000681*((100-x)^2) - 0.00032*(100-x)))
  return(y)  
}
ab.hab = function(x){
  y = (1-(0.00004678*((100-x)^2) + 0.000313*(100-x)))
  return(y)
}

# calculate effected habitat for each species

dat.sp$ASeffect.mix = other.sp(dat.sp$divert.mixed)
dat.sp$ASeffect.0 = other.sp(dat.sp$divert.0)
dat.sp$ASeffect.1 = other.sp(dat.sp$divert.1)

dat.sp$ABeffect.mix = ab.hab(dat.sp$divert.mixed)
dat.sp$ABeffect.0 = ab.hab(dat.sp$divert.0)
dat.sp$ABeffect.1 = ab.hab(dat.sp$divert.1)

```
### Grouping Variable = Irrigation Ditches
```{r Calculate irrigation ditch outputs}
addDitches = read_excel('~/Programming/Trutta/HSHEP/Emaui/fixNodesDitches.xlsx', sheet = 1)
addDitches = addDitches[ ,c(2,3)]
colnames(addDitches)[1]<- "BASINID"
ditch.dat = left_join(datw, addDitches, by = "BASINID")
View(ditch.dat)
div.mixed = nodes[, c(1:2)]
colnames(div.mixed)<- c("DSN1", "water.div.mix")
div.0 = nodes0[ ,c(1:2)]
colnames(div.0)<- c("DSN1", "water.div.0")
div.1 = nodes1[ ,c(1:2)]
colnames(div.1)<- c("DSN1", "water.div.1")

ditch.dat1 = left_join(ditch.dat, div.mixed, by = "DSN1") %>% 
  left_join(div.0, by = "DSN1") %>%
  left_join(div.1, by = "DSN1")
writexl::write_xlsx(ditch.dat1, paste(getwd(), "/eastMaui/rawData/ditchVolumes.xlsx", sep = ""))

```

### Grouping Variable = Watersheds
#### Habitat results
```{r Output habitat by watershed}
# output = summary of habitat by watershed and by species
WshedID = unique(scenario.test$Wshed)
WshedCount = c(1:n_distinct(WshedID))
summaries = tibble(WshedID, WshedCount)
species.abbvs = c("AB", "AS", "ES", "LC", "MG", "NG", "SH", "SS")

output = scenario.test %>%
  group_by(Wshed) %>%      #LINE to change GROUP ID var
  summarise_at(vars(c(4:27)), sum)

label = output[, 1]
All0 = round(output[ ,c(10:17)], 2)
# original = round(original, 2)
percentChangeMixed = ((output[ ,c(2:9)]- output[ ,c(10:17)])/output[ ,c(10:17)])*100
percentChangeAll1 = ((output[ ,c(18:25)]- output[ ,c(10:17)])/output[ ,c(10:17)])*100

scenario.changes = bind_cols(label,All0,percentChangeMixed, percentChangeAll1)
scenario.changes[is.na(scenario.changes)]<- 0

# *******************************************
# summaries by Wshed
Wshed = c(1:n_distinct(scenario.changes))
totalHab.All0 = rowSums(All0)
totalHab.All0[is.na(totalHab.All0)]<- 0
totalHab.mixed = rowSums(output[ ,c(2:9)])
totalHab.mixed[is.na(totalHab.mixed)]<- 0
totalHab.All1 = rowSums(output[ ,c(18:25)])
totalHab.All1[is.na(totalHab.All1)]<- 0
Wshed.out = data.frame(Wshed, totalHab.All0, totalHab.mixed, totalHab.All1)

Wshed.out$percentDelta.mixed = ((Wshed.out[, 3]-Wshed.out[ ,2])/Wshed.out[,2])*100
Wshed.out$percentDelta.mixed[is.na(Wshed.out$percentDelta.mixed)]<- 0
Wshed.out$percentDelta1 = ((Wshed.out[, 4]-Wshed.out[ ,2])/Wshed.out[,2])*100
Wshed.out$percentDelta1[is.na(Wshed.out$percentDelta1)]<- 0
Wshed.out$percentDelta.mixed[is.infinite(Wshed.out$percentDelta.mixed)]<- 0
Wshed.out$percentDelta1[is.infinite(Wshed.out$percentDelta1)]<- 0

Wshed.long = gather(Wshed.out, `percentDelta.mixed`, `percentDelta1`,  key = "NodeValue", value = "PercentChange")
Wshed.long$PercentChange[is.infinite(Wshed.long$PercentChange)]<- 0
Wshed.long$lPercentChange = log(Wshed.long$PercentChange)
Wshed.long$lPercentChange[is.infinite(Wshed.long$lPercentChange)]<- 0
Wshed.long$lPercentChange[is.infinite(Wshed.long$lPercentChange)]<- 0

ggplot(Wshed.out, aes(x = Wshed, y = (`totalHab.All0`))) +
  geom_bar(stat = 'identity') +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(y = "(All 0)") +
  theme_classic()+
  ggtitle("Watershed habitat summary - Baseline")

ggplot(Wshed.long, aes(x = Wshed, y = PercentChange, fill = NodeValue)) +
  geom_bar( stat = 'identity', position = 'dodge') +
  ylim(0, 1.2*max(Wshed.long$PercentChange)) +
  labs(y = "(PercentChange)") +
  theme_classic() +
  ggtitle("Watershed Habitat Changes under New Scenarios")

# generate summary table
colnames(Wshed.out)[2:6] = c("Total Habitat 0", "Total Habitat Mixed", "Total Habitat 1", "Percent Change Mixed", "Percent Change 1")

Wshed.out$`Percent Change Mixed`=percent(Wshed.out$`Percent Change Mixed`)

Wshed.out$`Percent Change 1`=percent(Wshed.out$`Percent Change 1`)

Wshed.tab = kable(Wshed.out, digits = 1, align = 'c',
                  caption = "Summary of Habitat Changes by Watershed")
Wshed.tab
```

**Space for interpretation of preceding plots and table**.

#### Water Diversion results - Water shunted to diversions grouped by watershed
```{r Output water by watershed}

# output = summary of water by watershed

outputw = scenario.testw %>%
  group_by(WshedID) %>%      #LINE to change GROUP ID var
  summarise_at(vars(mixed:All_1), sum)

# *******************************************
# calculate percent change from output
label = outputw[ ,1]
All0 = outputw[ ,3]
colnames(All0)[1] = "All0"
percentChangeMixed = ((outputw[ ,c(2)]- outputw[ ,c(3)])/outputw[ ,c(3)])*100
percentChangeAll1 = ((outputw[ ,c(4)]- outputw[ ,c(3)])/outputw[ ,c(3)])*100

scenario.changesw = bind_cols(label,All0,percentChangeMixed, percentChangeAll1)
scenario.changesw[is.na(scenario.changesw)]<- 0
colnames(scenario.changesw)[3]<- c("pChangeMixed")
colnames(scenario.changesw)[4]<- c("pChangeAll1")

# *******************************************
# summaries by watersheds
Wshed = c(1:n_distinct(scenario.changesw))
totalDiv.All0 = rowSums(All0)
totalDiv.All0[is.na(totalDiv.All0)]<- 0
totalDiv.mixed = rowSums(outputw[ ,c(2)])
totalDiv.mixed[is.na(totalDiv.mixed)]<- 0
totalDiv.All1 = rowSums(outputw[ ,c(4)])
totalDiv.All1[is.na(totalDiv.All1)]<- 0
Wshed.outw = data.frame(Wshed, totalDiv.All0, totalDiv.mixed, totalDiv.All1)

Wshed.outw$percentDelta.Mixed = ((Wshed.outw[, 3]-Wshed.outw[ ,2])/Wshed.outw[,2])*100
Wshed.outw$percentDelta.Mixed[is.na(Wshed.outw$percentDelta.Mixed)]<- 0
Wshed.outw$percentDelta1 = ((Wshed.outw[, 4]-Wshed.outw[ ,2])/Wshed.outw[,2])*100
Wshed.outw$percentDelta1[is.na(Wshed.outw$percentDelta1)]<- 0
Wshed.outw$percentDelta.Mixed[is.infinite(Wshed.outw$percentDelta.Mixed)]<- 0
Wshed.outw$percentDelta1[is.infinite(Wshed.outw$percentDelta1)]<- 0
# reshape data to plot percent delta

Wshed.outw.long = gather(Wshed.outw, `percentDelta.Mixed`, `percentDelta1`,  key = "NodeValue", value = "PercentChange")
Wshed.outw.long$PercentChange[is.infinite(Wshed.outw.long$PercentChange)]<- 0
Wshed.outw.long$lPercentChange = log(Wshed.outw.long$PercentChange)
Wshed.outw.long$lPercentChange[is.infinite(Wshed.outw.long$lPercentChange)]<- 0
Wshed.outw.long$lPercentChange[is.infinite(Wshed.outw.long$lPercentChange)]<- 0


# barplots
ggplot(Wshed.outw, aes(x = Wshed, y = (totalDiv.All0))) +
  geom_bar(stat = 'identity') +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(y = "All 0") +
  theme_classic()+
  ggtitle("Watershed Diversions - Baseline")

ggplot(Wshed.outw.long, aes(x = Wshed, y = PercentChange, fill = NodeValue)) +
  geom_bar( stat = 'identity', position = 'dodge') +
  ylim(0, 1.2*max(Wshed.outw.long$lPercentChange)) +
  labs(y = "PercentChange") +
  theme_classic() +
  ggtitle("Watershed Diversions - New Scenarios")

# summary table
colnames(Wshed.outw)[2:6] = c("Total Diverted - All 0", "Total Diverted - Mixed", "Total Diverted - All 1", "Percent Change - Mixed", "Percent Change - All 1")

Wshed.outw$`Percent Change - Mixed`= percent(Wshed.outw$`Percent Change - Mixed`)
Wshed.outw$`Percent Change - All 1`= percent(Wshed.outw$`Percent Change - All 1`)

Wshed.outw.tab = kable(Wshed.outw, digits = 1, align = 'c',
                       caption = "Summary of Water Diversions by Watershed")
Wshed.outw.tab

```


**Space for interpretation of preceding plots and table**.

#### Habitat results by species, grouped by Watershed
```{r Habitat output by species}

species.abbvs = c("AB", "AS", "ES", "LC", "MG", "NG", "SH", "SS")

AB = scenario.changes[ ,c(1,2,10,18)]
AS = scenario.changes[ ,c(1,3,11,19)]
ES = scenario.changes[ ,c(1,4,12,20)]
LC = scenario.changes[ ,c(1,5,13,21)]
MG = scenario.changes[ ,c(1,6,14,22)]
NG = scenario.changes[ ,c(1,7,15,23)]
SH = scenario.changes[ ,c(1,8,16,24)]
SS = scenario.changes[ ,c(1,9,17,25)]
species.list = list(AB,AS,ES,LC,MG,NG,SH,SS)
species.plots = list()

# list of barplots
for(i in 1:length(species.list)){
  colnames(species.list[[i]])<- c("Wshed", "All 0", "Mixed", "All 1")
  
  species.list[[i]][,5] = ((species.list[[i]][ ,2]-species.list[[i]][,3])/species.list[[i]][,3])*100
  species.list[[i]][,6] = ((species.list[[i]][ ,4]-species.list[[i]][,3])/species.list[[i]][,3])*100
  colnames(species.list[[i]])[c(5,6)]<-c("PercentChangeMixed", "PercentChangeAll1")
  species.list[[i]][is.na(species.list[[i]])]<- 0
  
  dat.long = gather(species.list[[i]], "PercentChangeMixed", "PercentChangeAll1", key = "NodeValue", value = "PercentChange")
  dat.long$lPercentChange = log(dat.long$PercentChange)
  dat.long$lPercentChange[is.na(dat.long$lPercentChange)]<- 0
  dat.long$lPercentChange[is.infinite(dat.long$lPercentChange)]<- 0
  
  
  g.sp = ggplot(dat.long, aes(x = Wshed, y = lPercentChange, fill = NodeValue )) +
    geom_bar( stat = 'identity', position = 'dodge') +
    theme_classic() +
    ggtitle(paste0("Species: ", species.abbvs[i], "; Grouped by: ", colnames(scenario.changes)[1], sep = " "))
  species.plots[[i]] <- g.sp

}

# list of tables
species.tabs = list()
for(i in 1:length(species.list)){
  species.list[[i]]$PercentChangeMixed = percent(species.list[[i]]$PercentChangeMixed)
  species.list[[i]]$PercentChangeAll1 = percent(species.list[[i]]$PercentChangeAll1)
  df = as.data.frame(species.list[[i]])
  df$Mixed[is.infinite(df$Mixed)]<- 0
  df$`All 1`[is.infinite(df$`All 1`)]<- 0
  # summary tables for each species
  IDvar.tab = kable(df, digits = 1, align = 'c')
  
  species.tabs[[i]]<- IDvar.tab
  
}

species.plots[[1]]
species.tabs[[1]]

species.plots[[2]]
species.tabs[[2]]

species.plots[[3]]
species.tabs[[3]]

species.plots[[4]]
species.tabs[[4]]

species.plots[[5]]
species.tabs[[5]]

species.plots[[6]]
species.tabs[[6]]

species.plots[[7]]
species.tabs[[7]]

species.plots[[8]]
species.tabs[[8]]

```

**Space for interpretation of preceding plots and tables**.
